---
title: "Getting Started"
author: '<h4>Authors: <i>`r auths <- eval(parse(text = gsub("person","c",read.dcf("../DESCRIPTION",
  fields = "Authors@R"))));paste(auths[names(auths)=="given"],auths[names(auths)=="family"],
  collapse = ", ")`</i></h4>'
date: '<h4>Vignette updated: <i>`r format( Sys.Date(), "%b-%d-%Y")`</i></h4>'
output:
  BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{echoAI}
    %\usepackage[utf8]{inputenc}
    %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=TRUE, message=FALSE}
library(echoAI)
```

```{r, echo=FALSE}
## Several examples query remote Zenodo/GitHub resources.
## Gate on internet access so R CMD check works offline.
has_internet <- function() {
    z <- try(suppressWarnings(
        readLines("https://github.com", n = 1L, warn = FALSE)
    ), silent = TRUE)
    !inherits(z, "try-error")
}
run_online <- has_internet()
```

# Introduction

`echoAI` provides API access to variant-level AI/ML predictions,
currently centred on three tools:

- **IMPACT** (Inference and Modeling of Phenotype-related ACtive Transcription)
  -- predicts transcription factor (TF) binding at motif sites by learning
  epigenomic profiles, primarily from [ENCODE](https://www.encodeproject.org/).
  The 707 annotations cover a wide range of immune and non-immune cell types,
  making IMPACT scores especially useful for prioritising causal variants in
  immune-mediated diseases. All IMPACT data are aligned to **hg19**.
  Tabix-indexed versions are hosted on Zenodo
  ([doi:10.5281/zenodo.7062238](https://doi.org/10.5281/zenodo.7062238))
  for rapid remote querying.

- **SpliceAI** -- predicts the probability that a variant alters mRNA splicing.
  Results can be obtained via a local VCF/TSV or the Broad Institute API.

- **Deep learning annotations** (Basenji, DeepSEA) -- query precomputed
  variant-level scores from deep learning models of chromatin accessibility
  and gene expression.

# IMPACT

## Query IMPACT annotations

The primary entry point is `IMPACT_query()`, which queries tabix-indexed
IMPACT annotation and LD-score files hosted on Zenodo.
This is the fastest way to annotate a small set of SNPs.

```{r query-impact, eval=FALSE}
## Prepare example query data
query_dat <- echodata::BST1[1:50,]

## Query IMPACT annotations (European population, wide format)
annot_wide <- IMPACT_query(
    query_dat = query_dat,
    types = "annot",
    populations = "EUR",
    output_format = "wide"
)
head(annot_wide)
```

You can also retrieve results in long format, which is useful for
downstream enrichment analysis. Setting `add_metadata = TRUE` appends
tissue/cell/TF information to each row.

```{r query-long, eval=FALSE}
annot_long <- IMPACT_query(
    query_dat = query_dat,
    types = "annot",
    populations = "EUR",
    output_format = "long",
    add_metadata = TRUE
)
head(annot_long)
```

## Annotation key

The annotation key maps each of the 707 IMPACT annotation IDs to its
source study, tissue, cell type/line, and transcription factor.

```{r annotation-key, eval=FALSE}
annot_key <- IMPACT_get_annotation_key(save_key = FALSE)
head(annot_key)
```

## Download full annotation matrices

For larger analyses (e.g. genome-wide or multi-locus), you can download
the full per-chromosome annotation matrices directly from the IMPACT
GitHub repository with `IMPACT_get_annotations()`.

**Note:** These files are large. Make sure you have sufficient memory
and storage.

```{r get-annotations, eval=FALSE}
## Get annotations for variants in the BST1 locus
BST1 <- echodata::BST1
annot_melt <- IMPACT_get_annotations(dat = BST1)
head(annot_melt)
```

To iterate across all chromosomes in a merged fine-mapping dataset:

```{r iterate-annotations, eval=FALSE}
merged_DT <- echodata::get_Nalls2019_merged()
ANNOT_MELT <- IMPACT_iterate_get_annotations(merged_DT = merged_DT)
```

## Post-processing

`IMPACT_postprocess_annotations()` converts the annotation table to
long format, identifies the top consensus SNP per locus, and adds a
combined cell-type label.

```{r postprocess, eval=FALSE}
ANNOT_MELT <- IMPACT_postprocess_annotations(ANNOT_MELT)
```

## Enrichment analysis

Enrichment is computed as the ratio of IMPACT signal in a given SNP group
(e.g. consensus, credible set, lead GWAS) to the proportion of SNPs in
that group. This tells you whether fine-mapped SNPs are enriched for
predicted TF binding relative to the background.

```{r enrichment, eval=FALSE}
## Compute enrichment for a single locus
enrich <- IMPACT_compute_enrichment(
    annot_melt = annot_melt,
    locus = "BST1"
)
head(enrich)

## Or iterate across all loci
ENRICH <- IMPACT_iterate_enrichment(ANNOT_MELT = ANNOT_MELT)
```

## Visualisation

### SNP group box plot

Compare IMPACT score distributions across SNP groups with
`IMPACT_snp_group_boxplot()`:

```{r boxplot, eval=FALSE}
bp <- IMPACT_snp_group_boxplot(
    TOP_IMPACT_all = TOP_IMPACT_all,
    snp_groups = c("GWAS lead", "UCS", "Consensus"),
    method = "wilcox.test"
)
```

### Enrichment plots

Visualise enrichment results with bar and violin plots:

```{r enrichment-plot, eval=FALSE}
plots <- IMPACT_plot_enrichment(ENRICH = ENRICH)
```

### Locus plot

Create a multi-panel locus plot showing GWAS results, fine-mapping
posterior probabilities, and per-tissue IMPACT scores:

```{r locus-plot, eval=FALSE}
impact_plot <- IMPACT_plot_impact_score(annot_melt = annot_melt)
```

### Heatmap

Generate a ComplexHeatmap of mean IMPACT scores across loci and SNP groups:

```{r heatmap, eval=FALSE}
mat_meta <- IMPACT_heatmap(ANNOT_MELT = ANNOT_MELT)
```

# SpliceAI

## Run SpliceAI

`SPLICEAI_run()` is the main entry point. It dispatches to the
appropriate backend (API, local TSV, or VCF) depending on your input.

```{r spliceai-run, eval=FALSE}
query_dat <- echodata::BST1[1:50,]
res <- SPLICEAI_run(query_dat = query_dat)
```

## Visualise splice probabilities

```{r spliceai-plot, eval=FALSE}
plt <- SPLICEAI_plot(query_dat = res)
```

# Deep learning annotations

## Query deep learning scores

`DEEPLEARNING_query()` retrieves precomputed variant-level scores from
deep learning models (e.g. Basenji, DeepSEA) via tabix-indexed files.

```{r dl-query, eval=FALSE}
query_dat <- echodata::BST1[1:50,]
dl_res <- DEEPLEARNING_query(query_dat = query_dat)
```

## Visualise scores

```{r dl-plot, eval=FALSE}
plt <- DEEPLEARNING_plot(dl_res)
```


# Session Info

<details>

```{r Session Info}
utils::sessionInfo()
```

</details>

<br>
